\documentclass{article}
\usepackage{iclr2026_conference,times}
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}

\input{../figures/latex_includes.tex}

\newcommand{\method}{ADRD}
\newcommand{\factorized}{\textsc{Factorized}}
\newcommand{\coupled}{\textsc{Coupled-Local}}
\newcommand{\micro}{\textsc{Micro-Sequential}}

\title{Budgeted Selective Dependency Routing for Discrete Diffusion Language Models\\A Synthetic Feasibility Study}

\author{}

\begin{document}
\maketitle

\begin{abstract}
Discrete diffusion language models offer attractive parallel decoding, but their quality can degrade when simultaneously predicted tokens have strong mutual dependencies. Recent work typically responds by choosing one dependency regime globally: stay factorized, introduce a stronger coupled output family, or move toward more sequential decoding. We study a narrower question: can richer dependency handling be allocated only to the minority of regions that appear risky? We present Adaptive Dependency Routing Diffusion (\method), a budgeted controller that routes unresolved regions to one of three decoding modes: factorized prediction, short-span coupled decoding, or tiny micro-sequential updates. The current implementation is intentionally modest: a synthetic discrete-diffusion scaffold with heuristic routing rather than a trained language model. Even in that setting, selective routing improves token accuracy from $0.6222$ to $0.6825$ and dependency accuracy from $0.3446$ to $0.6245$ while spending extra computation on only $10.89\%$ of tokens. On a longer-context stress test, \method\ improves dependency accuracy by $12.95$ points. At the same time, a simpler coupled-only surrogate slightly outperforms full \method\ on token accuracy under the relaxed budget, so our evidence supports feasibility rather than a definitive adaptive-routing win. The resulting paper is a controlled pilot study: selective dependency allocation looks promising, but stronger validation on real diffusion language models remains necessary.
\end{abstract}

\input{sections/introduction}
\input{sections/related_work}
\input{sections/method}
\input{sections/setup}
\input{sections/results}
\input{sections/limitations}
\input{sections/conclusion}

\bibliographystyle{iclr2026_conference}
\bibliography{references}

\end{document}
